Validation of Visual Statistical Inference, Applied to Linear Models

### 2012-07-03 paper_jasa.tex
cut out some of the plot signal strength discussion

This proves the statement above. We further see:
\[
E\left[ (1 - p_D)^{m-1}\right] =  \sum_{k=0}^{m-1} {m-1 \choose k} (-1)^k E[p_D^k] = 1-(m-1)E[p_D] + O(E[p_D^2]).
\]
This implies that as $p_D$, the $p$-value, becomes smaller, the probability that the observer identifies the data plot goes up and,
conversely, as $m$, the number of choices given in the lineup, goes up the probability to pick the data plot  goes down. 
}
%
%
%
%Let us think of a lineup as a head-to-head comparison of the test statistic and $m-1$ null plots. 
%%Let the data plot have a $p$-value of $p_D$, and the null plots $p$ values of $p_{0, i}$ with $1 \le i \le m-1$.
%We know that for each comparison the probability that the data, on which the plot is based, has the smaller $p$ value is 
%\begin{eqnarray*}
%P(p_D < p_{0,i}) &=& 1 - P(p_{i,0} \le p_D) = 1- \int_0^1  P(p_{i,0} \le p_D \mid p_D=t) f_{p_D}(t) dt =  \\
%&=& 1 - \int_0^1 F_{p_{0,i}}(t) f_{p_D}(t) dt = 1 - \int_0^1 t f_{p_D}(t) dt = 1 - E[p_D],
%\end{eqnarray*}
%which, in particular, is independent of $p_{0,i}$ for all $i$.
%
%Let us make the assumption that an observer is able to identify  the chart corresponding to the data with the smallest $p$-value. Further we will assume that all observers  have the same ability in identifying the data plot.
%
%With that, we define $Z$  as the number of null plots in a lineup, for which the $p$-value $p_{0,i}$ is smaller than $p_D$.  
%
%Then $Z \sim B_{m-1, E[p_D]}$, and the probability that an observer will pick the data plot in a given lineup is 
%\[
%P(Z=0) = \left(1 - E[p_D] \right)^{m-1} = P(p_D \le p_0), \ \ \ \text{ where } p_0 = \min_{1 \le i \le m-1}  \ p_{0,i}.
%\]
\end{proof}

@@

Now assume that the same lineup plot is evaluated by $K$ independent observers. %W.l.o.g.\green{Can we get rid of cliches like WLOG?} 
Let the last plot be the data plot and plots 1 through $m-1$ the null plots. Let $n_i$ be the number of observers who picked plot $i$ as the plot with the structure least consistent with the null hypothesis, corresponding to random variable $N_i$.  Then  $(N_1, N_2, ..., N_m) \sim$ Mult$_{\pi_1, ..., \pi_m}$ with $\sum_i \pi_i = 1$ and $\sum_{k=1}^{m} n_k = K$. We can estimate $\pi_i$ as $\widehat{\pi_i} = n_i/K$.

For the distribution of $N_m$, %the number of times out of $K$ that observers picked the last plot to show the structure least consistent with the null hypothesis, 
we get
\[
N_m\sim \left \{ 
\begin{array}{ll}
B_{K, \hh{E\left[ (1 - p_D)^{m-1}\right]}} & \text { under } H_1\\
B_{K, 1/m} & \text { under } H_0
\end{array}
\right .
\]

By matching the expected values, we therefore get an expression  for the $p$-value $p_D$ in the data 
%under $H_1$  %% no such thing aas $p$-value under H_1 - even though it's just an unfortunate choice of word.
as:
\begin{equation}\label{eqn:power_estimate}
%{E[p_D]} = 1- \left(\frac{E[N_m]}{K}\right)^{1/(m-1)}.
\hh{1 - E[N_m]/K = (m-1)E[p_D] + O(E[p_D^2])}
\end{equation}

Note that the left hand side in Equation \ref{eqn:power_estimate} is independent of the parameter $\theta$, but just based on the lineup evaluation by independent observers. While this allows us to compute this value independently of the parametric setting, it also means that observers might pick out the data plot for a reason not related to the value of $\theta$. 
\hh{Also note, that this equation yields an estimate of $p_D$ that is bounded by $1/(m-1)$. We should therefore not make the decision of significance depend on this estimate, but only use this equation after the fact. 
Both of these topics are further highlighted in the application section.}

@@

\green{Where is Figure \ref{fig:pval_definition} referred to and discussed?} \blue{Are we going to keep both figures \ref{fig:pval_definition} and \ref{fig:pval_plot_signal}?}

\green{Explain why the term ``plot signal'' is introduced}

@@

Note that $\hat{p}_D$ is bounded within $(0, 1/(m-1))$. This estimate has to be interpreted carefully, since the choice of a large lineup size, $m$, will make the value look `significant' in a traditional sense. Values close to the upper bound of $1/(m-1)$ should therefore be interpreted as `at least as high', rather than their actual numeric value.
%\hh{For $p$-values below $1/(m-1)$ visual $p$-values are very similar to actual $p$-values, the correlation for those values is: 0.36, 0.79, and -0.48 based on $n_1 = 44, n_2 = 45$, and $n_3= 26$ lineups, respectively.  }  \blue{This is not happening as we see in figure \ref{fig:pval_plot_signal}. But it is true for experiment 2, right?. }
%is equivalent to classical test $p$-value when $p$-value is less than or equal to $1/(m-1)$.  
%and beyond that the plot signal strength levels off on a point bigger than 0.05. Thus if plot signal strength is less than 0.05, we may decide to reject the null hypothesis.

@@

\begin{figure*}[hbtp]
   \centering
       \scalebox{0.5}{\includegraphics{p_val_plot_signal.pdf}}
       \caption{Conventional test $p$-value ($p_D$) vs plot signal strength ($\hat p_D$). The solid line is a 45 degree line which represents the equity shown in Equation \ref{eqn:power_estimate}. p-values are shown on square root scale scale.}
    %   \hh{Mahbub, another change: could you change the y label to 'estimate of visual p-value $\hat{p_D}$ ?}
       \label{fig:pval_plot_signal}
\end{figure*}

>>>>

This proves the statement above. 
%We further see:
%\[
%E\left[ (1 - p_D)^{m-1}\right] =  \sum_{k=0}^{m-1} {m-1 \choose k} (-1)^k E[p_D^k] = 1-(m-1)E[p_D] + O(E[p_D^2]).
%\]
This implies that as $p_D$, the $p$-value, becomes smaller, the probability that the observer identifies the data plot goes up and,
conversely, as $m$, the number of choices given in the lineup, goes up the probability to pick the data plot  goes down. 
}
%
%
%
%Let us think of a lineup as a head-to-head comparison of the test statistic and $m-1$ null plots. 
%%Let the data plot have a $p$-value of $p_D$, and the null plots $p$ values of $p_{0, i}$ with $1 \le i \le m-1$.
%We know that for each comparison the probability that the data, on which the plot is based, has the smaller $p$ value is 
%\begin{eqnarray*}
%P(p_D < p_{0,i}) &=& 1 - P(p_{i,0} \le p_D) = 1- \int_0^1  P(p_{i,0} \le p_D \mid p_D=t) f_{p_D}(t) dt =  \\
%&=& 1 - \int_0^1 F_{p_{0,i}}(t) f_{p_D}(t) dt = 1 - \int_0^1 t f_{p_D}(t) dt = 1 - E[p_D],
%\end{eqnarray*}
%which, in particular, is independent of $p_{0,i}$ for all $i$.
%
%Let us make the assumption that an observer is able to identify  the chart corresponding to the data with the smallest $p$-value. Further we will assume that all observers  have the same ability in identifying the data plot.
%
%With that, we define $Z$  as the number of null plots in a lineup, for which the $p$-value $p_{0,i}$ is smaller than $p_D$.  
%
%Then $Z \sim B_{m-1, E[p_D]}$, and the probability that an observer will pick the data plot in a given lineup is 
%\[
%P(Z=0) = \left(1 - E[p_D] \right)^{m-1} = P(p_D \le p_0), \ \ \ \text{ where } p_0 = \min_{1 \le i \le m-1}  \ p_{0,i}.
%\]
\end{proof}

@@

%Now assume that the same lineup plot is evaluated by $K$ independent observers. %W.l.o.g.\green{Can we get rid of cliches like WLOG?} 
%Let the last plot be the data plot and plots 1 through $m-1$ the null plots. Let $n_i$ be the number of observers who picked plot $i$ as the plot with the structure least consistent with the null hypothesis, corresponding to random variable $N_i$.  Then  $(N_1, N_2, ..., N_m) \sim$ Mult$_{\pi_1, ..., \pi_m}$ with $\sum_i \pi_i = 1$ and $\sum_{k=1}^{m} n_k = K$. We can estimate $\pi_i$ as $\widehat{\pi_i} = n_i/K$. 
%
%For the distribution of $N_m$, %the number of times out of $K$ that observers picked the last plot to show the structure least consistent with the null hypothesis, 
%we get
%\[
%N_m\sim \left \{ 
%\begin{array}{ll}
%B_{K, \hh{E\left[ (1 - p_D)^{m-1}\right]}} & \text { under } H_1\\
%B_{K, 1/m} & \text { under } H_0
%\end{array}
%\right .
%\]
%
%By matching the expected values, we therefore get an expression  for the $p$-value $p_D$ in the data 
%%under $H_1$  %% no such thing aas $p$-value under H_1 - even though it's just an unfortunate choice of word.
%as:
%\begin{equation}\label{eqn:power_estimate}
%%{E[p_D]} = 1- \left(\frac{E[N_m]}{K}\right)^{1/(m-1)}.
%\hh{1 - E[N_m]/K = (m-1)E[p_D] + O(E[p_D^2])}
%\end{equation}
%
%Note that the left hand side in Equation \ref{eqn:power_estimate} is independent of the parameter $\theta$, but just based on the lineup evaluation by independent observers. While this allows us to compute this value independently of the parametric setting, it also means that observers might pick out the data plot for a reason not related to the value of $\theta$. 
%\hh{Also note, that this equation yields an estimate of $p_D$ that is bounded by $1/(m-1)$. We should therefore not make the decision of significance depend on this estimate, but only use this equation after the fact. 
%Both of these topics are further highlighted in the application section.}

@@

\green{Where is Figure \ref{fig:pval_definition} referred to and discussed?}
% \blue{Are we going to keep both figures \ref{fig:pval_definition} and \ref{fig:pval_plot_signal}?}
%
%\green{Explain why the term ``plot signal'' is introduced}

@@

%Note that $\hat{p}_D$ is bounded within $(0, 1/(m-1))$. This estimate has to be interpreted carefully, since the choice of a large lineup size, $m$, will make the value look `significant' in a traditional sense. Values close to the upper bound of $1/(m-1)$ should therefore be interpreted as `at least as high', rather than their actual numeric value.
%\hh{For $p$-values below $1/(m-1)$ visual $p$-values are very similar to actual $p$-values, the correlation for those values is: 0.36, 0.79, and -0.48 based on $n_1 = 44, n_2 = 45$, and $n_3= 26$ lineups, respectively.  }  \blue{This is not happening as we see in figure \ref{fig:pval_plot_signal}. But it is true for experiment 2, right?. }
%is equivalent to classical test $p$-value when $p$-value is less than or equal to $1/(m-1)$.  
%and beyond that the plot signal strength levels off on a point bigger than 0.05. Thus if plot signal strength is less than 0.05, we may decide to reject the null hypothesis.

@@

%\begin{figure*}[hbtp]
%   \centering
%       \scalebox{0.5}{\includegraphics{p_val_plot_signal.pdf}}
%       \caption{Conventional test $p$-value ($p_D$) vs plot signal strength ($\hat p_D$). The solid line is a 45 degree line which represents the equity shown in Equation \ref{eqn:power_estimate}. p-values are shown on square root scale scale.}
%    %   \hh{Mahbub, another change: could you change the y label to 'estimate of visual p-value $\hat{p_D}$ ?}
%       \label{fig:pval_plot_signal}
%\end{figure*}
